Growth of the South African Economy:\\An Empirical Evaluation

### 2014-08-14 Articles/Article_EcolEcon_2014/Paper2.Rnw
standardTriPlot & standardScatterPlot now give colors.

<<data, include=FALSE, eval=TRUE>>=
# Load historical data
historicalData <- rbind.fill(REXS1960, IST, Leeds) # From the EconData package
# Load coefficients from all models in one large data frame
REXS1960_Coeffs <- loadPostProcessedData(Source = "REXS1960", kind = "coeffs")
IST_Coeffs <- loadPostProcessedData(Source = "IST", kind = "coeffs")
Leeds_Coeffs <- loadPostProcessedData(Source = "Leeds", kind = "coeffs")
AllCoeffs <- rbind(REXS1960_Coeffs, IST_Coeffs, Leeds_Coeffs)
AllCoeffs$Source <- as.factor(replace(AllCoeffs$Source, which(AllCoeffs$Source=="REXS1960"), "REXS"))
AllCoeffs$Source <- relevelFactor(AllCoeffs$Source, levs = c("REXS", "IST", "Leeds"))
AllCoeffs$energy <- relevelFactor(AllCoeffs$energy, levs = energyLevels)
# Load fitted data from all models in one large data frame
REXS1960_Fitted <- loadPostProcessedData(Source = "REXS1960", kind = "fitted")
IST_Fitted <- loadPostProcessedData(Source = "IST", kind = "fitted")
Leeds_Fitted <- loadPostProcessedData(Source = "Leeds", kind = "fitted")
AllFitted <- rbind(REXS1960_Fitted, IST_Fitted, Leeds_Fitted)
AllFitted$Source <- replace(AllFitted$Source, which(AllFitted$Source=="REXS1960"), "REXS")
AllFitted$Source <- relevelFactor(as.factor(AllFitted$Source), levs = c("REXS", "IST", "Leeds"))
AllFitted$energy <- relevelFactor(AllFitted$energy, levs=energyLevels)
@

@@

<<Historical_Data_1, fig.pos="H", fig.align='center', fig.width=6.6, fig.height=3, fig.cap="Historical data.", eval=TRUE>>=
mdata <- melt(historicalData, 
              measure.vars = c("iGDP", "iK", "iL", "iXp", "iXf", "iU"), 
              variable.name = "variable",
              value.name = "value")
historicalPlot(mdata, 
               facet_formula = Country ~ Source, 
               lines_formula = variable + value ~ Year, 
               line_types = c(1,2,3,4,5,6)) + 
  xlab("") + 
  ylab("Indexed (1960=1 or 1978=1)")
@

@@

<<Historical_Data_2, fig.pos="H", fig.align='center', fig.width=8, fig.height=3, fig.cap="Historical data.", eval=TRUE>>=
historicalPlot(mdata, 
               facet_formula=Country~variable, 
               lines_formula=Source+value~Year, 
               line_types=c(1,2,3)) +  
  xlab("") + 
  ylab("Indexed (1960=1 or 1978=1)")
@

@@

<<CD_lambda_theta_graph, fig.pos="H", fig.align='center', fig.width=4, fig.height=4.5, fig.cap="$\\lambda$ and $\\theta$ values for UK Cobb-Douglas resample models.">>=
cdCoeffsData <- subset(AllCoeffs, subset = model=="cd")
standardScatterPlot(data=cdCoeffsData, mapping=aes(scale, lambda), facet_formula=energy~Source, alpha=0.05) +
    scale_x_continuous(limits=c(0.95, 1.05), breaks=c(0.95, 1.0)) +
    scale_y_continuous(breaks=c(0.01, 0.015, 0.02)) +    
    labs(x=expression(theta), y=expression(lambda))
@

<<CD_Ternary_Scatter_Graph, fig.pos="H", fig.align='center', fig.width=3.9, fig.height=4, fig.cap="UK Cobb-Douglas resample model parameters.">>=
standardTriPlot(data=cdCoeffsData, facet_formula=energy~Source)
@

<<CD_GDP_Spaghetti_Graph, fig.pos="H", fig.align='center', fig.width=4, fig.height=4.5, fig.cap="UK Cobb-Douglas fitted models.">>=
cdFittedData <- subset(AllFitted, subset = model=="cd")
spaghettiPlot(data=cdFittedData, facet_formula=energy~Source, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

@@

<<LINEX_abg_traj_graph_REXS_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for REXS with $u$.">>=
triPlot(data=fortify(REXS1960_Models$REXS1960$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

@@

<<LINEX_abg_traj_graph_IST_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for IST with $u$.">>=
triPlot(data=fortify(IST_Models$IST$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

@@

<<LINEX_abg_traj_graph_Leeds_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for Leeds with $u$.">>=
triPlot(data=fortify(Leeds_Models$Leeds$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

>>>>

<<data, include=FALSE, eval=TRUE>>=
# Load historical data
historicalData <- rbind.fill(REXS1960, IST, Leeds) # From the EconData package

# The working directory is set to the directory that contains this file.
# To reach the post-processed data, we need to move 
# up two levels and into the data_postprocessed directory.
# So, in each of the loadPostProcessedData calls below, we use
# dir = "../../data_postprocessed".
pp_path <- file.path("..", "..", "data_postprocessed")

# Load coefficients from all models in one large data frame
REXS1960_Coeffs <- loadPostProcessedData(Source = "REXS1960", kind = "coeffs",dir = pp_path)
IST_Coeffs <- loadPostProcessedData(Source = "IST", kind = "coeffs",dir = pp_path)
Leeds_Coeffs <- loadPostProcessedData(Source = "Leeds", kind = "coeffs", dir = pp_path)
AllCoeffs <- rbind(REXS1960_Coeffs, IST_Coeffs, Leeds_Coeffs)
AllCoeffs$Source <- as.factor(replace(AllCoeffs$Source, which(AllCoeffs$Source=="REXS1960"), "REXS"))
AllCoeffs$Source <- relevelFactor(AllCoeffs$Source, levs = c("REXS", "IST", "Leeds"))
AllCoeffs$energy <- relevelFactor(AllCoeffs$energy, levs = energyLevels)
# Load fitted data from all models in one large data frame
REXS1960_Fitted <- loadPostProcessedData(Source = "REXS1960", kind = "fitted", dir = pp_path)
IST_Fitted <- loadPostProcessedData(Source = "IST", kind = "fitted", dir = pp_path)
Leeds_Fitted <- loadPostProcessedData(Source = "Leeds", kind = "fitted", dir = pp_path)
AllFitted <- rbind(REXS1960_Fitted, IST_Fitted, Leeds_Fitted)
AllFitted$Source <- replace(AllFitted$Source, which(AllFitted$Source=="REXS1960"), "REXS")
AllFitted$Source <- relevelFactor(as.factor(AllFitted$Source), levs = c("REXS", "IST", "Leeds"))
AllFitted$energy <- relevelFactor(AllFitted$energy, levs=energyLevels)
@

@@

<<Historical_Data_1, fig.pos="H", fig.align='center', fig.width=6.6, fig.height=6, fig.cap="Historical data.", eval=TRUE>>=
mdata <- melt(historicalData, 
              measure.vars = c("iGDP", "iK", "iL", "iXp", "iXf", "iU"), 
              variable.name = "variable",
              value.name = "value")
historicalPlot(mdata, 
               mapping=aes(x=Year, y=value, group=variable, linetype=variable), 
               facet_formula = Country ~ Source, 
               line_types = c(1,5,2,6,4,3)) + 
  xlab("") + 
  ylab("Indexed (1960=1 or 1978=1)")
@

@@

<<Historical_Data_2, fig.pos="H", fig.align='center', fig.width=8, fig.height=6, fig.cap="Historical data.", eval=TRUE>>=
historicalPlot(mdata, 
               mapping=aes(x=Year, y=value, group=Source, linetype=Source), 
               facet_formula = Country ~ variable, 
               line_types = c(1,5,3)) + 
  xlab("") + 
  ylab("Indexed (1960=1 or 1978=1)")
@

@@

<<CD_lambda_theta_graph, fig.pos="H", fig.align='center', fig.width=6, fig.height=4.5, fig.cap="$\\lambda$ and $\\theta$ values for Cobb-Douglas resample models.">>=
cdCoeffsData <- subset(AllCoeffs, subset = model=="cd")
standardScatterPlot(data=cdCoeffsData, mapping=aes(x=scale, y=lambda, color=Source), 
                    facet_formula=country~energy, alpha=0.05) +
    scale_x_continuous(limits=c(0.8, 1.2), breaks=c(0.9, 1.0, 1.1)) +
    scale_y_continuous(limits=c(-0.1, 0.2), breaks=c(0.0, 0.1, 0.2)) +    
    labs(x=expression(theta), y=expression(lambda)) +
    guides(colour=guide_legend(override.aes = aes(alpha=1))) #+
#     geom_density2d()
@

<<CD_Ternary_Scatter_Graph, fig.pos="H", fig.align='center', fig.width=6, fig.height=4, fig.cap="Cobb-Douglas resample model parameters.">>=
standardTriPlot2(data=cdCoeffsData, 
                mapping=aes(color=Source),
                facet_formula=country~energy) +
    guides(colour=guide_legend(override.aes = aes(alpha=1)))
@

<<CD_GDP_Spaghetti_Graphs_US, fig.pos="H", fig.align='center', fig.width=4, fig.height=2.5, fig.cap="US Cobb-Douglas fitted models.">>=
cdFittedDataUS <- subset(AllFitted, subset = model=="cd" & Country=="US")
spaghettiPlot(data=cdFittedDataUS, facet_formula=Source~energy, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

<<CD_GDP_Spaghetti_Graphs_UK, fig.pos="H", fig.align='center', fig.width=4, fig.height=4.5, fig.cap="UK Cobb-Douglas fitted models.">>=
cdFittedDataUK <- subset(AllFitted, subset = model=="cd" & Country=="UK")
spaghettiPlot(data=cdFittedDataUK, facet_formula=Source~energy, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

<<CD_GDP_Spaghetti_Graphs_JP, fig.pos="H", fig.align='center', fig.width=4, fig.height=2.5, fig.cap="JP Cobb-Douglas fitted models.">>=
cdFittedDataJP <- subset(AllFitted, subset = model=="cd" & Country=="JP")
spaghettiPlot(data=cdFittedDataJP, facet_formula=Source~energy, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

<<CD_GDP_Spaghetti_Graphs_AT, fig.pos="H", fig.align='center', fig.width=5, fig.height=3.5, fig.cap="AT Cobb-Douglas fitted models.">>=
cdFittedDataAT <- subset(AllFitted, subset = model=="cd" & Country=="AT")
spaghettiPlot(data=cdFittedDataAT, facet_formula=Source~energy, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

<<CD_GDP_Spaghetti_Graphs_CN, fig.pos="H", fig.align='center', fig.width=4, fig.height=2.5, fig.cap="CN Cobb-Douglas fitted models.">>=
cdFittedDataCN <- subset(AllFitted, subset = model=="cd" & Country=="CN")
spaghettiPlot(data=cdFittedDataCN, facet_formula=Source~energy, level=spaghettiGraphLevel) +
    labs(y="Indexed GDP (1960=1)") +
  scale_x_continuous(breaks=c(1960, 1980, 2000))
@

@@

<<LINEX_abg_traj_graph_REXS_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for REXS with $u$.">>=
REXS1960_Models <- loadPostProcessedData(Source = "REXS1960", kind = "models", dir=pp_path)
triPlot(data=fortify(REXS1960_Models$REXS1960$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

@@

<<LINEX_abg_traj_graph_IST_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for IST with $u$.">>=
IST_Models <- loadPostProcessedData(Source = "IST", kind = "models", dir=pp_path)
triPlot(data=fortify(IST_Models$IST$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

@@

<<LINEX_abg_traj_graph_Leeds_UK_U, fig.pos="H", fig.align='center', fig.width=3, fig.height=3.5, fig.cap="UK LINEX trajectory for Leeds with $u$.">>=
Leeds_Models <- loadPostProcessedData(Source = "Leeds", kind = "models", dir=pp_path)
triPlot(data=fortify(Leeds_Models$Leeds$UK$linex$`iK+iL+iU`), grid_lines=5, x=alpha, y=beta, z=gamma, geom=geom_line, aes_string="colour = iYear + 1960")
@

### 2014-08-14 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
standardTriPlot & standardScatterPlot now give colors.

# Set working directory to the top level of the repository: Econ-Growt-R-Analysis
# This is the same directory from which the RStudio project runs.
# Setting this working directory eliminates the need to use dir="../.." in the 
# loadPostPorcessedData function calls below.
setwd(file.path("..", ".."))

# Load coefficients from all models in one large data frame
coeffs_df <- loadPostProcessedData(Source = "Calvin", kind = "coeffs")
# Load fitted data from all models in one large data frame
fitted_df <- loadPostProcessedData(Source = "Calvin", kind = "fitted")
# Load the original models.
oModels <- loadPostProcessedData(Source = "Calvin", kind = "models")
@

@@

Many who accept that energy use is closely correlated with
economic performance, regardless of the direction of causality,
express concern regarding the negative effects of 
rising energy prices and fossil fuel depletion 
on current and future economic growth.
\citet{Benes:2012tm}, incorporating views on both
geological physical constraints and 
technological solutions to oil production, 
suggest a near doubling of the real price of over the next decade.  
\citet{Kumhof:2012ua} point out that 
significant impacts on GDP growth could occur
for larger oil supply shocks, 
assuming production functions where oil plays a prominent role 
and reduced possibilities for substitution away from oil.  
\citet{Nel:2010fv} argue for the 
``\dots vital importance of energy security \dots'' 
on the basis of a model based on an explicit 
energy-based formulation of economic growth.  
\citet{Hamilton:2013vc} argues that 
``[c]oping with a final peak in world oil production \dots 
[will] have much in common with previous historical episodes that resulted from 
temporary geopolitical conflict, 
being associated with significant declines in employment and output. 
If the future decades look like the last 5 years, we are in for a rough time.''

The current trends in energy production and consumption and 
the possible dire impacts on the world economy 
as argued by the above-mentioned scholars 
raise the question of
whether and how to include appropriate energy indicators 
in economic growth modeling.  
Neoclassical economic growth theory attributes output~($Y$) to factors of production:
capital stock~($K$) and labor~($L$). 
The Cobb-Douglas production function 
is the canonical representation of indexed economic output ($y$) as a function of indexed 
labor~($l$), indexed capital stock~($k$), and an additional exogenous, time-dependent 
term.%
  \footnote{The time-dependent term in production functions is often linked
  to ``technological progress,'' but that is a very narrow reading of Solow's original intent:
  \begin{quote}
    It will be seen that I am using the phrase ``technical change''
    as a short-hand expression for any kind of shift 
    in the production function. 
    Thus slowdowns, speed-ups, improvements in the education of the labor force, 
    and all sorts of things will appear as ``technical change.''~\citep{solow1957technical}
  \end{quote}
  }
This production function can be expressed as
%
\begin{equation} \label{eq:CD}
  y = \theta \mathrm{e}^{\lambda(t-t_0)} k^\alpha l^\beta \;,
\end{equation}
%
\noindent where 
$y \equiv Y/Y_{0}$,
$\theta$ is a scale parameter,
e is the base of the natural logarithm, 
$\lambda$ is represents the pace of technological progress,
$t$ (time) is measured in years,
$k \equiv K/K_{0}$, 
$l \equiv L/L_{0}$, 
$Y$ is represented by GDP, 
$K$ is expressed in currency units, 
$L$ is expressed in workers or work-hours/year, and
0 subscripts indicate values at an initial year.%
\footnote{Dimensionless, indexed quantities are represented by 
lower-case symbols 
($y$, $k$, $l$, $e$, $q$, $x$, and $u$), and dimensional 
quantities are represented by upper-case symbols 
($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$). 
Model parameters are represented by Greek letters
($\alpha$, $\beta$, $\lambda$, $\theta$).}

Neoclassical economic growth theory as expressed in the Cobb-Douglas production 
function makes no attempt, beyond growth in capital~($k$) and labor~($l$), 
to describe the mechanisms that affect economic growth, 
nor does it attempt, beyond different fitting parameters,
to describe how such phenomena as
slowdowns, speedups, and improvements of the education of the labor force
might vary from one economy to the next. 
The parameters $\alpha$ and $\beta$ indicate the relative roles 
of capital and labor in the economy.
Typically, the constraint $\alpha + \beta = 1$ is assumed, 
indicating constant returns to scale.

@@

The remainder of this paper is organized as follows. 
Section~\ref{sec:lit_review} provides a necessarily brief overview of relevant literature.
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output and factors of production. 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion_of_Results} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes. 
An appendix provides details of all statistical results.

@@

\citet[p.~26]{Stern:2011bq} provides an overview of approaches to including energy 
in ecological economic and resource economic models of growth.  
\citeauthor{Stern:2011bq} points out that 
energy is seen as the \emph{central driver of growth} 
in ecological economic (including biophysical economic) models
and that energy \emph{is included} in resource economic models, 
but \emph{remains isolated} within the sub-discipline.  
To break through this theoretical impasse, 
Stern proposes a synthesized Constant Elasticity of Substitution (CES) 
production function 
where the constraints or abundance of energy 
determines whether to include energy explicitly 
as a factor of production.  
On the basis of this model, 
\citet[p. 45]{Stern:2011bq} points out that 
energy availability plays
``\dots a key role in enabling growth,''
despite the evidence of decreasing energy consumed per unit of output.  
He concludes, based on theory and time-series results, 
that including energy in the production function is warranted, 
and that further clarification is needed 
on the role of energy in economic growth.

The inclusion of energy as a factor of production 
in Cobb-Douglas and CES production functions 
is generally accepted by most 
resource and energy economists
\citep{Smulders:2003wc, vanderZwaan:2002wn, Kemfert:1998vd}.
Researchers interested in explaining long-term economic growth 
and possible constraints on growth 
have started to include energy as an explicit factor of production
\citep{Froling:2009vo, Tahvonen:2001tz}.  
An alternative approach, attempting to be 
``\dots more consistent with physical laws \dots'' \citep{Ayres:2001tx}
and classified by \citet{Stern:2011bq} as an ecological economic model of growth, 
is the LINEX model, 
wherein energy is the only factor of production,
and capital~($k$), labor~($l$), and energy~($e$)
combine to provide efficiencies that augment energy.
Technological change is manifest in these efficiencies,
as more useful work is done per joule of exergy \citep{Ayres:2005kn}.

@@

But, the near-exclusive focus on developed economies closes off 
the possibility of discovering any role that energy plays 
in the economic development process. 
To gain insights from different countries 
across different stages of development, 
we present a comparative analysis among representatives of
the World Bank’s main categories of countries, 
namely Developed Countries (DCs), 
Newly Industrialized Countries (NICs), 
OPEC countries, and 
Least Developed Countries (LDCs). 
Table~\ref{tab:Economy_Examples} shows our example countries.

@@

Figure~\ref{fig:Historical_Graph} presents historical data, 
including indexed values for 
output ($y$), 
capital stock ($k$), 
labor ($l$), and 
energy ($q$) 
for 1980--2011 (developed economies) and 1991--2011 (other economies). 
% Factors of production track each other closely, 
% and high covariances among these variables can be seen in 
% Tables~\ref{tab:Covariance_US}--\ref{tab:Covariance_ZM}. 
% \authNote{Do we want to discuss the covariances?}
%
<<Historical_Graph, fig.pos="H", fig.align='center', fig.width=3, fig.height=7.5, fig.cap="Historical data: indexed GDP ($y$), capital stock ($k$), labor ($l$), and thermal energy ($q$).">>=
mdata <- melt(historicalData, 
              measure.vars = c("iGDP", "iK", "iL", "iQp"), 
              value.name = "value", 
              variable.name = "variable") 
historicalPlot(mdata, 
               facet_formula = Country ~ Source, 
               lines_formula = variable + value ~ Year, 
               line_types = c(1,2,3,4)) +
  xlab("") +
  ylab("Indexed (1980-1 or 1991=1")
@

@@

Figure~\ref{fig:CD_Ternary_Scatter_Graph} shows ternary plots 
for $\alpha$, $\beta$, and $\gamma$ parameter values 
for Cobb-Douglas (without and with energy) resample models.
Each reample is shown as a gray dot on the graph. 
The fit to historical data is shown as a gray crosshairs.

@@

\twocolumn{}
<<CD_Ternary_Scatter_Graph, fig.pos="H", fig.align='center', fig.width=2.9, fig.height=7.7, fig.cap="Cobb-Douglas resample model parameters.">>=
cdCoeffsData <- filter(coeffs_df, model=="cd", nestStr %in% c("iK+iL", "iK+iL+iQp"))
standardTriPlot(data=cdCoeffsData, facet_formula=country~energy)
@

>>>>

# The working directory is set to the directory that contains this file.
# To reach the post-processed data, we need to move 
# up two levels and into the data_postprocessed directory.
# So, in each of the loadPostProcessedData calls below, we use
# dir = "../../data_postprocessed".
pp_path <- file.path("..", "..", "data_postprocessed")

# Load coefficients from all models in one large data frame
coeffs_df <- loadPostProcessedData(Source = "Calvin", kind = "coeffs", dir = pp_path)
# Load fitted data from all models in one large data frame
fitted_df <- loadPostProcessedData(Source = "Calvin", kind = "fitted", dir = pp_path)
# Load the original models.
oModels <- loadPostProcessedData(Source = "Calvin", kind = "models", dir = pp_path)
@

@@

Many who accept that energy use is closely correlated with
economic performance, regardless of the direction of causality,
express concern regarding the negative effects of 
rising energy prices and fossil fuel depletion 
on current and future economic growth.
\citet{Benes:2012tm}, incorporating views on both
geological physical constraints and 
technological solutions to oil production, 
suggest a near doubling of the real price of over the next decade.  
\citet{Kumhof:2012ua} point out that 
significant impacts on GDP growth could occur
for larger oil supply shocks, 
assuming production functions where oil plays a prominent role 
and reduced possibilities for substitution away from oil.  
\citet{Nel:2010fv} argue for the 
``\dots{}vital importance of energy security\dots'' 
on the basis of an economic growth model with explicit 
energy-based formulation.  
\citet{Hamilton:2013vc} argues that 
``[c]oping with a final peak in world oil production\dots%
[will] have much in common with previous historical episodes that resulted from 
temporary geopolitical conflict, 
being associated with significant declines in employment and output. 
If the future decades look like the last 5 years, we are in for a rough time.''

The current trends in energy production and consumption and 
the possible dire impacts on the world economy 
as argued by the above-mentioned scholars 
raise the question of
whether and how to include appropriate energy indicators 
in economic growth modeling.  
Neoclassical economic growth theory attributes output~($Y$) to factors of production:
capital stock~($K$) and labor~($L$). 
The Cobb-Douglas production function 
is the canonical representation of indexed economic output ($y$) as a function of indexed 
labor~($l$), indexed capital stock~($k$), and an additional exogenous, time-dependent 
term.%
  \footnote{The time-dependent term in production functions is often associated with
  ``technological progress,'' but that is a very narrow reading of Solow's original intent:
  \begin{quote}
    It will be seen that I am using the phrase ``technical change''
    as a short-hand expression for any kind of shift 
    in the production function. 
    Thus slowdowns, speed-ups, improvements in the education of the labor force, 
    and all sorts of things will appear as ``technical change.''~\citep{solow1957technical}
  \end{quote}
  }
This production function can be expressed as
%
\begin{equation} \label{eq:CD}
  y = \theta \mathrm{e}^{\lambda(t-t_0)} k^\alpha l^\beta \;,
\end{equation}
%
\noindent where 
$y \equiv Y/Y_{0}$,
$\theta$ is a scale parameter,
e is the base of the natural logarithm, 
$\lambda$ is represents the pace of technological progress,
$t$ (time) is measured in years,
$k \equiv K/K_{0}$, 
$l \equiv L/L_{0}$, 
$Y$ is represented by GDP, 
$K$ is expressed in currency units, 
$L$ is expressed in workers or work-hours/year, and
0 subscripts indicate values at an initial year.
The parameters $\alpha$ and $\beta$ indicate the relative roles 
of capital and labor in the economy.
Typically, the constraint $\alpha + \beta = 1$ is assumed, 
indicating constant returns to scale.%
  \footnote{Dimensionless, indexed quantities are represented by 
  lower-case symbols 
  ($y$, $k$, $l$, and $q$), and dimensional 
  quantities are represented by upper-case symbols 
  ($Y$, $K$, $L$, and $Q$). 
  Model parameters are represented by Greek letters
  ($\alpha$, $\beta$, $\lambda$, $\theta$).}

Neoclassical economic growth theory as expressed in the Cobb-Douglas production 
function makes no attempt, beyond growth in capital~($k$) and labor~($l$), 
to describe the mechanisms that affect economic growth, 
nor does it attempt, beyond different fitting parameters,
to describe how such phenomena as
slowdowns, speedups, and improvements of the education of the labor force
might vary from one economy to the next.

@@

The remainder of this paper is organized as follows. 
Section~\ref{sec:lit_review} provides a necessarily brief overview of relevant literature.
Section~\ref{sec:Coordinates} discusses the coordinates of analysis for the present study. 
Section~\ref{sec:Sources_of_Data} details sources of data and calculations to 
generate time series for economic output ($y$) and factors of production ($k$, $l$, and $q$). 
The procedures for and results from parameter estimation for the various production functions 
are discussed in Section~\ref{sec:Parameter_Estimation}. 
Section~\ref{sec:Discussion_of_Results} discusses implications of the parameter estimation 
results, and Section~\ref{sec:Conclusion} concludes.

@@

\citet[p.~26]{Stern:2011bq} provides an overview of approaches to including energy 
in ecological economic and resource economic models of growth.  
\citeauthor{Stern:2011bq} points out that 
energy is seen as the \emph{central driver of growth} 
in ecological economic (including biophysical economic) models
and that energy \emph{is included in} resource economic models, 
but \emph{remains isolated} within the sub-discipline.  
To break through this theoretical impasse, 
Stern proposes a synthesized Constant Elasticity of Substitution (CES) 
production function 
where the constraints or abundance of energy 
determines whether to include energy explicitly 
as a factor of production.  
On the basis of this model, 
\citet[p. 45]{Stern:2011bq} points out that 
energy availability plays
``\dots{}a key role in enabling growth,''
despite evidence of decreasing energy consumed per unit of output.  
He concludes, based on theory and time-series results, 
that including energy in the production function is warranted, 
and that further clarification is needed 
on the role of energy in economic growth.

The inclusion of energy as a factor of production 
in Cobb-Douglas and CES production functions 
is generally accepted by most 
resource and energy economists
\citep{Smulders:2003wc, vanderZwaan:2002wn, Kemfert:1998vd}.
Researchers interested in explaining long-term economic growth 
and possible constraints on growth 
have started to include energy as an explicit factor of production
\citep{Froling:2009vo, Tahvonen:2001tz}.  
An alternative approach, attempting to be 
``\dots more consistent with physical laws \dots'' \citep{Ayres:2001tx}
and classified by \citet{Stern:2011bq} as an ecological economic model of growth, 
is the LINEX model, 
wherein energy is the only factor of production,
and capital~($k$), labor~($l$), and energy~($q$)
combine to provide efficiencies that augment energy.
Technological change is manifest as increased efficiency:
more useful work done per joule of exergy \citep{Ayres:2005kn}.
There is no Solow-style term ($\mathrm{e}^{\lambda t}$) for technological change.

@@

But, the near-exclusive focus on developed economies closes off 
the possibility of discovering any role that energy plays 
in the economic development process. 
To gain insights from different countries 
across different stages of development, 
we present a comparative analysis among representatives of
the World Bank's main categories of countries, 
namely Developed Countries (DCs), 
Newly Industrialized Countries (NICs), 
OPEC countries, and 
Least Developed Countries (LDCs). 
Table~\ref{tab:Economy_Examples} shows our example countries.

@@

Figure~\ref{fig:Historical_Graph} presents historical data, 
including indexed values for 
output ($y$), 
capital stock ($k$), 
labor ($l$), and 
energy ($q$) 
for 1980--2011 (developed economies) and 1991--2011 (other economies). 
% Factors of production track each other closely, 
% and high covariances among these variables can be seen in 
% Tables~\ref{tab:Covariance_US}--\ref{tab:Covariance_ZM}. 
% \authNote{Do we want to discuss the covariances?}
%
<<Historical_Graph, fig.pos="H", fig.align='center', fig.width=3, fig.height=7.5, fig.cap="Historical data: indexed GDP ($y$), capital stock ($k$), labor ($l$), and thermal energy ($q$).">>=
mdata <- melt(historicalData, 
              measure.vars = c("iGDP", "iK", "iL", "iQp"), 
              value.name = "value", 
              variable.name = "variable") 
historicalPlot(data=mdata, 
               mapping=aes(x=Year, y=value, group=variable, linetype=variable), 
               facet_formula = Country ~ Source, 
               line_types = c(1,5,2,3)) +
  xlab("") +
  ylab("Indexed (1980=1 or 1991=1)")
@

@@

Figure~\ref{fig:CD_lambda_theta_graph} shows $\lambda$ and $\theta$ values
for Cobb-Douglas resample models with and without energy.
Each reample is shown as a gray dot on the graph. 
The fit to historical data is shown as a gray crosshairs.

Figure~\ref{fig:CD_Ternary_Scatter_Graph} shows ternary plots 
for $\alpha$, $\beta$, and $\gamma$ parameter values 
for Cobb-Douglas (without and with energy) resample models.

@@

<<CD_lambda_theta_graph, fig.pos="H", fig.align='center', fig.width=2.8, fig.height=7.0, fig.cap="$\\lambda$ and $\\theta$ values for CD resample models.">>=
cdCoeffsData <- filter(coeffs_df, model=="cd", nestStr %in% c("iK+iL", "iK+iL+iQp"))
standardScatterPlot(data=cdCoeffsData, mapping=aes(x=scale, y=lambda), facet_formula=country~energy, alpha=0.05) +
    scale_x_continuous(limits=c(0.8, 1.2), breaks=c(0.8, 1.0, 1.2)) +
    scale_y_continuous(limits=c(-0.05, 0.15), breaks=c(0, 0.05, 0.1)) +    
    labs(x=expression(theta), y=expression(lambda))
@

\twocolumn{}
<<CD_Ternary_Scatter_Graph, fig.pos="H", fig.align='center', fig.width=2.9, fig.height=7.7, fig.cap="Cobb-Douglas resample model parameters.">>=
standardTriPlot(data=cdCoeffsData, facet_formula=country~energy)
@

### 2014-08-14 Articles/Article_Heun_Warr_2014/EcolEcon_HeunWarr.Rnw
standardTriPlot & standardScatterPlot now give colors.

<<data, include=FALSE, eval=TRUE>>=
# Load historical data
historicalData <- REXS # From the EconData package
# Load coefficients from all models in one large data frame
coeffs_df <- tbl_df(REXS_Coeffs)
# Load fitted data from all models in one large data frame
fitted_df <- tbl_df(REXS_Fitted)

# Load the original models.
oModels <- REXS_Models
@

>>>>

<<data, include=FALSE, eval=TRUE>>=
# Load historical data
historicalData <- REXS # From the EconData package

# The working directory is set to the directory that contains this file.
# To reach the post-processed data, we need to move 
# up two levels and into the data_postprocessed directory.
# So, in each of the loadPostProcessedData calls below, we use
# dir = "../../data_postprocessed".
pp_path <- file.path("..", "..", "data_postprocessed")

# Load coefficients from all models in one large data frame
coeffs_df <- loadPostProcessedData(Source = "REXS", kind = "coeffs", dir=pp_path)
# Load fitted data from all models in one large data frame
fitted_df <- loadPostProcessedData(Source = "REXS", kind = "fitted", dir=pp_path)
# Load the original models.
oModels <- loadPostProcessedData(Source = "REXS", kind = "models", dir=pp_path)
@

### 2014-08-14 Articles/Article_JAE_2014/ZAGrowth_JAE.Rnw
standardTriPlot & standardScatterPlot now give colors.

# Load coefficients from all models in one large data frame
coeffs_df <- loadPostProcessedData(Source = "Calvin", kind = "coeffs", dir = "../../data_postprocessed")
coeffs_df_ZA <- subset(coeffs_df, subset=country=="ZA")
# Load fitted data from all models in one large data frame
fitted_df <- loadPostProcessedData(Source = "Calvin", kind = "fitted", dir = "../../data_postprocessed")
fitted_df_ZA <- subset(fitted_df, subset=Country=="ZA")
# Load the original models.
oModels <- loadPostProcessedData(Source = "Calvin", kind = "models", dir = "../../data_postprocessed")
oModels_ZA <- oModels$Calvin$ZA
@

>>>>

# The working directory is set to the directory that contains this file.
# To reach the post-processed data, we need to move 
# up two levels and into the data_postprocessed directory.
# So, in each of the loadPostProcessedData calls below, we use
# dir = "../../data_postprocessed".
pp_path <- file.path("..", "..", "data_postprocessed")

# Load coefficients from all models in one large data frame
coeffs_df <- loadPostProcessedData(Source = "Calvin", kind = "coeffs", dir=pp_path)
coeffs_df_ZA <- subset(coeffs_df, subset=country=="ZA")
# Load fitted data from all models in one large data frame
fitted_df <- loadPostProcessedData(Source = "Calvin", kind = "fitted", dir=pp_path)
fitted_df_ZA <- subset(fitted_df, subset=Country=="ZA")
# Load the original models.
oModels <- loadPostProcessedData(Source = "Calvin", kind = "models", dir=pp_path)
oModels_ZA <- oModels$Calvin$ZA
@
